---
title: "DS311 Project Q2"
author: "Gabrielle Salamanca"
date: "March 17, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## We-R-Finished

Before we can dive into the question, we must load in the necessary libraries and read the dataset into the markdown.

```{r}
# loading libraries
library(readxl)
library(caret)
library(ggplot2)
library(dplyr)

# dataset overview
salary <- read_excel("C:/Users/knigh/OneDrive/Desktop/Github/We-R-Finished/salary/salary_data_states.xlsx")
```

Next, let's rename the columns to make it easier for us to read.

```{r}
# renaming
colnames(salary)[1] = "Case Number"
colnames(salary)[2] = "Case Status"
colnames(salary)[3] = "Received Date"
colnames(salary)[4] = "Decision Date"
colnames(salary)[5] = "Employer Name"
colnames(salary)[6] = "Submitted Prevail Wage"
colnames(salary)[7] = "SPrW Unit"
colnames(salary)[8] = "Submitted Paid Wage"
colnames(salary)[9] = "SPaW Unit"
colnames(salary)[10] = "Job Title"
colnames(salary)[11] = "Work City"
colnames(salary)[12] = "Required Edu"
colnames(salary)[13] = "Required College Major"
colnames(salary)[14] = "Exp Req"
colnames(salary)[15] = "Exp Req (Months)"
colnames(salary)[16] = "Citizenship"
colnames(salary)[17] = "Prevail Wage SOC Code"
colnames(salary)[18] = "PWSOC Title"
colnames(salary)[19] = "Work State"
colnames(salary)[20] = "WS Abb"
colnames(salary)[21] = "WPostal Code"
colnames(salary)[22] = "Full Time"
colnames(salary)[23] = "Visa Class"
colnames(salary)[24] = "Prevail Wage/Yr"
colnames(salary)[25] = "Paid Wage/Yr"
colnames(salary)[26] = "Job Title Sub"
colnames(salary)[27] = "Order"
```

Now, we must remove the sub job titles that have teaching jobs. The question specifically asks for data-related salaries in the job sub-categories, therefore we must remove: professor, attorney, assistant professor, and teacher.

```{r}
# filter
sal <- salary %>%
  filter(!grepl("professor", `Job Title Sub`, ignore.case = TRUE) & 
           !grepl("attorney", `Job Title Sub`, ignore.case = TRUE) &
           !grepl("assistant professor", `Job Title Sub`, ignore.case = TRUE) & 
           !grepl("teacher", `Job Title Sub`, ignore.case = TRUE))
jobTitleSub <- as.factor(sal$`Job Title Sub`)
summary(jobTitleSub)
```

### II. What states (of those I am willing to move to) have the highest paying data-related salaries?

This is quite a broad range of states, but we will remove any US territories first and foremost for obvious reasons. This means Guam, Guamam, Palau, Northern Mariana Islands, Puerto Rico, and Virgin Islands will not be considered. 

```{r}
# filtering states
datsal <- sal %>%
  filter(!grepl("Guam", `Work State`, ignore.case = TRUE) & 
           !grepl("Guamam", `Work State`, ignore.case = TRUE) &
           !grepl("Palau", `Work State`, ignore.case = TRUE) & 
           !grepl("Northern Mariana Islands", `Work State`, ignore.case = TRUE) &
           !grepl("Puerto Rico", `Work State`, ignore.case = TRUE) &
           !grepl("Virgin Islands", `Work State`, ignore.case = TRUE))
```

Now, that still leaves us with a large amount of states. When I tried running graphs to include work state and the job title sub variables, R had difficulty running them, because it was far too many to handle. 

Now, the question asks *"What states (of those I am willing to move to)"*. Considering the group is staying in the Bay Area, we may be more picky about where we would want to live. California is much more friendly towards immigrants and quite open to people of all walks of life. We may be biased to stay within the West Coast, especially considering the natural disasters and how certain minorities are treated here either by law or general perception. This would greatly narrow the scope, so I have decided to use the top 10 work states as my pool, minus California. My reasoning is because we are already in California, and the state has a frequency of:

```{r}
CAfreq <- table(datsal$`Work State`)["California"]
CAfreq
```

This state completely dwarfs the others. See here:

```{r}
TXfreq <- table(datsal$`Work State`)["Texas"]
NYfreq <- table(datsal$`Work State`)["New York"]
NJfreq <- table(datsal$`Work State`)["New Jersey"]
ILfreq <- table(datsal$`Work State`)["Illinois"]
MAfreq <- table(datsal$`Work State`)["Massachusetts"]
VAfreq <- table(datsal$`Work State`)["Virginia"]
PAfreq <- table(datsal$`Work State`)["Pennsylvania"]
WAfreq <- table(datsal$`Work State`)["Washington"]
MIfreq <- table(datsal$`Work State`)["Michigan"]
NCfreq <- table(datsal$`Work State`)["North Carolina"]
# dataframe
Frequency <- c(TXfreq, NYfreq, NJfreq, ILfreq, MAfreq, VAfreq, PAfreq, WAfreq, MIfreq, NCfreq)
stateFreq <- data.frame(Frequency)
stateFreq
```

California would skew the graphs if it were included, so we will only use these states for this question.

```{r}
topState <- c("Texas", "New York",
          	"New Jersey", "Illinois", "Massachussetts",
          	"Virginia", "Pennsylvania", "Washington",
          	"Michigan", "North Coralina")
ssal <- datsal[datsal$`Work State` %in% topState, ]
```

#### a. Differences between job sub-categories?

```{r}
# variables
stateAbb <- ssal$`WS Abb`
jobSub <- ssal$`Job Title Sub`

# job sub/state plot
ggplot(ssal, aes(x = stateAbb, fill = jobSub)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Work State Abbreviation", y = "Count", title = "Job Subcategories in Each State")

# state/job sub
ggplot(ssal, aes(x = jobSub, fill = stateAbb)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Job Title Sub", y = "Count", title = "States for Each Job Subcategory")
```

We see that software engineer and business analyst are the largest proportions compared to any other job sub-category in the first graph, especially in the second. Let's see how each sub-type does in terms of salary.

```{r}
# variable(s)
paidWage <- ssal$`Paid Wage/Yr`

# paid wage boxplot
ggplot(ssal, aes(x = jobSub, y = paidWage, fill = jobSub)) +
  geom_boxplot() +
  labs(title = "Paid Wage/Year by Job Sub-Type",
   	x = "Job Sub-Type",
   	y = "Paid Wage/Year",
   	fill = "Job Sub-Type")

# no outliers
ggplot(ssal, aes(x = jobSub, y = paidWage, fill = jobSub)) +
  geom_boxplot() +
  labs(title = "Paid Wage/Year by Job Sub-Type",
   	x = "Job Sub-Type",
   	y = "Paid Wage/Year",
   	fill = "Job Sub-Type") +
  scale_y_continuous(limits = quantile(paidWage, c(0.1, 0.9)))
```

We see that we have many outliers in the first graph, especially in software engineer. But it seems management consultant and data scientist have larger salary ranges compared to the rest of the job sub-types. Now, if we did get rid of the data points outside of 10th and 90th percentile, we get pretty much a similar story. An analyst makes less than a data scientist, management consultant, and software engineer. These three jobs pay well, but it seems software engineer is the only one that's being advertised to the most out of the them. Let's see if it holds true for each state. We will only showcase the graphs without the outliers.

```{r}
# variable
ssal$Sub <- factor(ssal$`Job Title Sub`, levels = c("business analyst", "data analyst", "data scientist", "management consultant", "software engineer"))

# Texas
ssal %>%
  filter(`Work State` == "Texas", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "darkorange1", color = "darkorange4") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in Texas")

# New York
ssal %>%
  filter(`Work State` == "New York", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "lightblue", color = "skyblue4") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in New York")

# New Jersey
ssal %>%
  filter(`Work State` == "New Jersey", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "moccasin", color = "khaki3") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in New Jersey")

# Illinois
ssal %>%
  filter(`Work State` == "Illinois", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "darkorange1", color = "navyblue") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in Illinois")

# Massachussetts
ssal %>%
  filter(`Work State` == "Massachussetts") %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "goldenrod2", color = "darkslateblue") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in Massachussetts")

# Virginia
ssal %>%
  filter(`Work State` == "Virginia", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "seashell", color = "dodgerblue4") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in Virginia")

# Pennsylvania
ssal %>%
  filter(`Work State` == "Pennsylvania", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "gold2", color = "royalblue4") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in Pennsylvania")

# Washington
ssal %>%
  filter(`Work State` == "Washington", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "springgreen4", color = "gold3") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution Washington")

# Michigan
ssal %>%
  filter(`Work State` == "Michigan", `Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "darkslateblue", color = "goldenrod3") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution Michigan")

# North Carolina
ssal %>%
  filter(`Work State` == "North Carolina") %>%
  ggplot(aes(x = Sub, y = `Paid Wage/Yr`)) +
  geom_boxplot(fill = "skyblue", color = "steelblue") +
  labs(x = "Job Sub-Type", y = "Paid Wage/Year", title = "Salary Distribution in North Carolina")
```


#### b. Which companies have the highest salaries for those sub-types?

```{r}
employer <- as.factor(ssal$`Employer Name`)
summary(employer)
paidWage <- ssal$`Paid Wage/Yr`
```


#### c. Will the answer change if I take standard of living into account?

```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE}
```