---
title: "DS311 Data Explore"
author: "Gabrielle Salamanca"
date: "March 14, 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## We-R-Finished Project

### 1. Loading necessary libraries & reading dataset into R

```{r}
# loading libraries
library(readxl)
library(caret)
library(ggplot2)
library(dplyr)

# dataset overview
salary <- read_excel("C:/Users/knigh/OneDrive/Desktop/Github/We-R-Finished/salary/salary_data_states.xlsx")
dim(salary)
```

### 2. Renaming columns for sake of ease

```{r}
# renaming
colnames(salary)[1] = "Case Number"
colnames(salary)[2] = "Case Status"
colnames(salary)[3] = "Received Date"
colnames(salary)[4] = "Decision Date"
colnames(salary)[5] = "Employer Name"
colnames(salary)[6] = "Submitted Prevail Wage"
colnames(salary)[7] = "SPrW Unit"
colnames(salary)[8] = "Submitted Paid Wage"
colnames(salary)[9] = "SPaW Unit"
colnames(salary)[10] = "Job Title"
colnames(salary)[11] = "Work City"
colnames(salary)[12] = "Required Edu"
colnames(salary)[13] = "Required College Major"
colnames(salary)[14] = "Exp Req"
colnames(salary)[15] = "Exp Req (Months)"
colnames(salary)[16] = "Citizenship"
colnames(salary)[17] = "Prevail Wage SOC Code"
colnames(salary)[18] = "PWSOC Title"
colnames(salary)[19] = "Work State"
colnames(salary)[20] = "WS Abb"
colnames(salary)[21] = "WPostal Code"
colnames(salary)[22] = "Full Time"
colnames(salary)[23] = "Visa Class"
colnames(salary)[24] = "Prevail Wage/Yr"
colnames(salary)[25] = "Paid Wage/Yr"
colnames(salary)[26] = "Job Title Sub"
colnames(salary)[27] = "Order"
names(salary)
```

### 3. What are the job title subcategories & their numbers

```{r}
jobTitleSub <- as.factor(salary$`Job Title Sub`)
summary(jobTitleSub)
# dataframe
Sub_Job <- c("Software Engineer", "Business Analyst", "Assistant Prof", "Teacher",
         	"Data Scientist", "Data Analyst", "Attorney", "Management Consultant")
Frequency <- c(99364, 27811, 18866, 13912, 1227, 3840, 1488, 770)
job2 <- data.frame(Sub_Job, Frequency)
print(job2)
```

Was expecting higher number of data scientist and analyst
No expecting assistant professor & teacher in top 5

```{r}
city <- as.factor(salary$`Work City`)
summary(city)
top10 <- c("New York", "Mountain View", "San Francisco", "San Diego", "San Jose", "Houston", "Santa Clara", "Atlanta", "Chicago", "Dallas")
freq <- c(6714, 6712, 6458, 2983, 2925, 2957, 2684, 2388, 2351, 2277)
topCity <- data.frame(top10, freq)
print(topCity)
```

Not surprised NY and CA areas are in top 5 cities
Expected LA to be in top
wait i didn't expect to see my city, but it sure is a lot of CA areas

### 4. Who are the top 10 employers

```{r}
employName <- as.factor(salary$`Employer Name`)
summary(employName)
top10 <- c("Google Inc.", "Fujitsu America, Inc.", "Intel Corporation", 
           "Microsoft Corporation", "Qualcomm Technologies, Inc.", "Facebook, Inc.", 
           "Cerner Corporation", "Cisco Systems, Inc.", "Hitachi Consulting Corporation", "Dallas Independent School District")
freq <- c(6213, 1814, 1781, 1364, 1277, 1210, 1038, 1011, 812, 2277)
topEmploy <- data.frame(top10, freq)
print(topEmploy)
```

We know Google, Intel, Microsoft, Facebook, Cisco. The others I don't recognize. Also Dallas Independent School District is one of the top 10 employers?

### 5. Top 10 states to work in

```{r}
state <- as.factor(salary$`Work State`)
summary(state)
top10 <- c("California", "Texas", "New York", "New Jersey", "Illinois",
           "Massachussets", "Virgina", "Pennsylvania", "Washington", "Michigan")
freq <- c(46782, 15498, 11373, 10198, 7411, 6848, 6031, 4725, 4610, 3844)
topState <- data.frame(top10, freq)
print(topState)
```

as expected CA is the largest, but WOW did the numbers really vary after. I was expecting TX and NY, but def not NJ. 

### 6. Top 10 Jobs

```{r}
jobTitle <- as.factor(salary$`Job Title`)
summary(jobTitle)
top10 <- c("Software Engineer", "Business Analyst", "Sr Software Engineer", "Assistant Professor", "Computer Software Engineer, Apps",
           "Teacher", "Data Analyst", "Sr Business Analyst", "Staff Software Engineer", "Elementary Bilungual Teacher")
freq <- c(58221, 19516, 13847, 10134, 1808, 1566, 1524, 1437, 1335, 1020)
topJob <- data.frame(top10, freq)
print(topJob)
```

Dataset separated the ones with different cases, had to add up a few. Wasn't expecting teachers, esp the last one. engineer and analyst are popular jobs it seems

```{r}
paidWageYear <- salary$`Paid Wage/Yr`
summary(paidWageYear)
# plots
plot(paidWageYear, jobTitleSub)
plot(paidWageYear, jobTitle)
```

Norm seems to be under $500,000 wage/year. If only the table can tell me which sub jobs they were... I will need to revisit the jobTitle one, but the norm still holds.

### 7. Subset time

```{r}
salary1 <- salary[,-c(6,7,9)]
salary2 <- salary1[,-c(9:15)]
salary3 <- salary2[, -c(10:15)]
salary4 <- salary3[,-c(1:4)]
```

Easier to handle the dataset, but also in need to cut down any unnecessary columns. Also doing in increments, because it didn't like how I tried doing range with the first one.

### 8. Trying the plots again

```{r}
plot(jobTitleSub, paidWageYear)
```


```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE}
```