---
title: "DS311 Q6"
author: "Gabrielle Salamanca"
date: "April 20, 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## We-R-Finished

```{r}
library(readr)
library(readxl)
library(car)
library(dplyr)
library(ggplot2)
library(scales)
library(gridExtra)
```

### Question 6

When looking over at the three given questions, there was a common sub-question:

**Will the answer change if I take standard of living into account?**

While it leans more hypothetical, it did bring thoughts of perhaps we need to quantify it to answer it. In this question, we're going to dive a bit deeper on that question using the Cost of Living and the salary dataset.

### What is standard/cost of living?

From https://worldpopulationreview.com/state-rankings/cost-of-living-index-by-state:

"Cost of living refers to the amount needed to cover basic expenses, such as food, shelter, transportation, and healthcare."

Of course, each state will vary in their cost of living. From what we know, California is one of the most expensive states to live in. Let's take the shelter expense for example. We know that housing prices have risen to insane numbers in this state. For example, in San Mateo, a single family home can cost upwards to 2 million USD. If one goes to Hillsborough, known to be an expensive area, is double that. Rent is also extremely high, a one bedroom apartment can range between 2 thousand to 4 thousand USD. 

There could be many reasons for this such as the population density, tax rates, etc. When moving out of California, or at least the highly populated areas of the state, the costs will go down. Let's say perhaps somewhere in the Midwest or the South. Housing prices are more likely to be less than 1 million USD, therefore the cost of living index may lower.

### Why is it important?

In our particular project, having data on the cost of living for each region helps employers determine the wage of the job. Of course, certain *genre* of jobs will have higher pay than others, and specific *types* of jobs within that genre will have better wage than the others. When we take the salary dataset into account, we can use this data to find out if one would be able to live in a state with the salary they will receive. Of course, I will only use the top 10 states for data-related jobs I found in Q2.

```{r}
# reading salary dataset
salary <- read_excel("C:/Users/knigh/OneDrive/Desktop/Github/We-R-Finished/salary/salary_data_states.xlsx")
colnames(salary)[19] = "Work State"
colnames(salary)[26] = "Job Title Sub"
# filtering
sal <- salary %>%
  filter(!grepl("professor", `Job Title Sub`, ignore.case = TRUE) &
       	!grepl("attorney", `Job Title Sub`, ignore.case = TRUE) &
       	!grepl("assistant professor", `Job Title Sub`, ignore.case = TRUE) &
       	!grepl("teacher", `Job Title Sub`, ignore.case = TRUE))
datsal <- sal %>%
  filter(!grepl("Guam", `Work State`, ignore.case = TRUE) &
       	!grepl("Guamam", `Work State`, ignore.case = TRUE) &
       	!grepl("Palau", `Work State`, ignore.case = TRUE) &
       	!grepl("Northern Mariana Islands", `Work State`, ignore.case = TRUE) &
       	!grepl("Puerto Rico", `Work State`, ignore.case = TRUE) &
       	!grepl("Virgin Islands", `Work State`, ignore.case = TRUE))
```

### About the dataset

The Cost of Living dataset is from the same site I have mentioned in the previous question, which was found by Russell Chan.

```{r}
# reading CoL dataset
COL <- read_csv("C:/Users/knigh/OneDrive/Desktop/Github/We-R-Finished/CostOfLiving2023.csv", show_col_types = FALSE)
coldf <- COL[, !(names(COL) %in% c("fips"))]
head(coldf)
```

Within this dataset, there are 50 entries, one for each state in the USA. It contains data on the population, the growth, and various cost indexes for each state. The main columns we are interested in are the cost and the state, because we are planning to merge the CoL and salary dataset.

### Prepping the dataset

We first create a subset in coldf with only the state and costIndex columns before merging it with the salary dataset.

```{r}
# subset
costate <- subset(coldf, select = c(state, costIndex))
# merging
salcost <- merge(datsal, costate,  by.x = "Work State", by.y = "state")
head(salcost)
```

Next, we will make a new column to create an adjusted wage, where we divide the paid wage per year by the cost index to even the field. We will also have a specific subset that only includes: employer name, job title sub, work state, paid wage per year, adjusted year, and Cost index.

```{r}
# adjusted wage
salcost$ADJUSTED_WAGE <- (salcost$PAID_WAGE_PER_YEAR/salcost$costIndex)*100
# new subset
salcost <- salcost[, c("EMPLOYER_NAME", "Job Title Sub", "Work State", "WORK_STATE_ABBREVIATION", "PAID_WAGE_PER_YEAR", "ADJUSTED_WAGE", "costIndex")]

# for each state, median
allAdjustMed <- aggregate(salcost$ADJUSTED_WAGE, by = list(salcost$`Work State`), FUN = median)
names(allAdjustMed) <- c("State", "Median Adjusted Wage")
allPaidMed <- aggregate(salcost$PAID_WAGE_PER_YEAR, by = list(salcost$`Work State`), FUN = median)
names(allPaidMed) <- c("State", "Median Paid Wage")

# sorting
allAdjustMed <- allAdjustMed[order(-allAdjustMed$'Median Adjusted Wage'),]
allPaidMed <- allPaidMed[order(-allPaidMed$'Median Paid Wage'),]

# rename columns
colnames(salcost)[1] = "Employer"
colnames(salcost)[4] = "State Abb"
colnames(salcost)[5] = "Paid Wage/Yr"
colnames(salcost)[6] = "Adjusted Wage"
colnames(salcost)[7] = "Cost Index"
head(salcost)
```

### The Question

What I aim to answer with the merged dataset is this:

What are the states with the largest change in salary adjusted wage? which are the top and bottom states in terms of salary considering adjusted wage? If we look at job subtitles, how does salary change for top and bottom 5 considering the cost of living?

Previously in Q2, I have used the average and max to answer part b and . Cost of living dataset does use averages to then make a baseline, and I was planning to use averages to answer this question. But, by Ana's suggestion, median would be far more stable than using the mean. The mean tends to be skewed if there's more or less of something, and knowing that software engineer has major outliers, this made much more sense.

```{r}
salcost$Sub <- factor(salcost$`Job Title Sub`, levels = c("business analyst", "data analyst", "data scientist", "management consultant", "software engineer"))
salcost$'Median Salary' <- median(salcost$`Adjusted Wage`)
# topState <- c("Texas", "New York",
#     		 "New Jersey", "Illinois", "Massachusetts",
#     		 "Virginia", "Pennsylvania", "Washington",
#     		 "Michigan", "North Carolina")
# salcost <- salcost[salcost$`Work State` %in% topState, ]

# mean
salcost %>%
  filter(`Adjusted Wage` > quantile(`Adjusted Wage`, 0.05) & 
           `Adjusted Wage` < quantile(`Adjusted Wage`, 0.95)) %>%
  group_by(Sub, `Employer`) %>%
  summarise(mean_salary = mean(`Adjusted Wage`)) %>%
  arrange(Sub, desc(mean_salary)) %>%
  group_by(Sub) %>%
  top_n(n = 5, wt = mean_salary)

# median
salcost %>%
  filter(`Adjusted Wage` > quantile(`Adjusted Wage`, 0.05) & 
           `Adjusted Wage` < quantile(`Adjusted Wage`, 0.95)) %>%
  group_by(Sub, `Employer`) %>%
  summarise(med_salary = median(`Adjusted Wage`)) %>%
  arrange(Sub, desc(med_salary)) %>%
  group_by(Sub) %>%
  top_n(n = 5, wt = med_salary)
```

```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE}
```
