---
title: "Q6"
author: "Ana Paula Felix de Queiroz"
output:
  word_document: default
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{=html}
<style type="text/css">

h1.title {
  font-size: 38px;
  font-family: "Times New Roman", Times, serif;
  color: #2a6592;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: #2a6592;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/pauli/OneDrive/Documentos/Classes 2023/311_tech")

library(car)

```

## We-R-Finished

```{r}
library(readr)
library(readxl)
library(dplyr)
```

### Question 6

When looking over at the three given questions, there was a common sub-question:

**Will the answer change if I take standard of living into account?**

While it leans more hypothetical, it did bring thoughts of perhaps we need to quantify it to answer it. In this question, we're going to dive a bit deeper on that question using the Cost of Living and the salary data set.

### What is standard/cost of living?

From <https://worldpopulationreview.com/state-rankings/cost-of-living-index-by-state>:

"Cost of living refers to the amount needed to cover basic expenses, such as food, shelter, transportation, and healthcare."

Of course, each state will vary in their cost of living. From what we know, California is one of the most expensive states to live in. Let's take the shelter expense for example. We know that housing prices have risen to insane numbers in this state. For example, in San Mateo, a single family home can cost upwards to 2 million USD. If one goes to Hillsborough, known to be an expensive area, is double that. Rent is also extremely high, a one bedroom apartment can range between 2 thousand to 4 thousand USD.

There could be many reasons for this such as the population density, tax rates, etc. When moving out of California, or at least the highly populated areas of the state, the costs will go down. Let's say perhaps somewhere in the Midwest or the South. Housing prices are more likely to be less than 1 million USD, therefore the cost of living index may lower.

### Why is it important?

In our particular project, having data on the cost of living for each region helps employers determine the wage of the job. Of course, certain *genre* of jobs will have higher pay than others, and specific *types* of jobs within that genre will have better wage than the others. When we take the salary dataset into account, we can use this data to find out if one would be able to live in a state with the salary they will receive. Of course, I will only use the top 10 states for data-related jobs I found in Q2.

```{r}
# reading salary dataset
salary <- read_excel("salary_data_states.xlsx")
#names(salary)
#dim(salary)
colnames(salary)[19] = "Work_State"
colnames(salary)[26] = "Job_Title_Sub"
# filtering
sal <- salary %>%
  filter(!grepl("professor", `Job_Title_Sub`, ignore.case = TRUE) &
  		 !grepl("attorney", `Job_Title_Sub`, ignore.case = TRUE) &
  		 !grepl("assistant professor", `Job_Title_Sub`, ignore.case = TRUE) &
  		 !grepl("teacher", `Job_Title_Sub`, ignore.case = TRUE))
datsal <- sal %>%
  filter(!grepl("Guam", `Work_State`, ignore.case = TRUE) &
  		 !grepl("Guamam", `Work_State`, ignore.case = TRUE) &
  		 !grepl("Palau", `Work_State`, ignore.case = TRUE) &
  		 !grepl("Northern Mariana Islands", `Work_State`, ignore.case = TRUE) &
  		 !grepl("Puerto Rico", `Work_State`, ignore.case = TRUE) &
  		 !grepl("Virgin Islands", `Work_State`, ignore.case = TRUE))
```

### About the data set

The Cost of Living data set is from the same site I have mentioned in the previous question, which was found by Russell Chan.

```{r}
# reading CoL dataset
COL <- read_csv("CostOfLiving2023.csv", show_col_types = FALSE)
coldf <- COL[, !(names(COL) %in% c("fips"))]
head(coldf)
```

Within this dataset, there are 50 entries, one for each state in the USA. It contains data on the population, the growth, and various cost indexes for each state. The main columns we are interested in are the cost and the state, because we are planning to merge the CoL and salary dataset.

### Prepping the dataset

We first create a subset in coldf with only the state and costIndex columns before merging it with the salary dataset.

```{r}
# subset
costate <- subset(coldf, select = c(state, costIndex))
# merging
salcost <- merge(datsal, costate,  by.x = "Work_State", by.y = "state")
#head(salcost)

```

Next, we will make a new column to create an adjusted wage, where we divide the paid wage per year by the cost index to even the field. We will also have a specific subset that only includes: employer name, job title sub, work state, paid wage per year, adjusted year, and Cost index.

```{r}
salcost$ADJUSTED_WAGE <- (salcost$PAID_WAGE_PER_YEAR / salcost$costIndex)*100

# new subset
salcost <- salcost[, c("EMPLOYER_NAME", "Job_Title_Sub", "Work_State", "PAID_WAGE_PER_YEAR", "ADJUSTED_WAGE", "costIndex", "WORK_STATE_ABBREVIATION")]


#unique(salcost$Job_Title_Sub)

```

### The Question

-   **Which are the top and bottom states in terms of salary and what change when we consider adjusted wage.**

Here we will look at the top 5 states median wage, and see how much the wages change when taking into consideration the cost of living in these countries.

Top 5 states:

| Top 5 States  | Paid Wage | Adjusted Wage |
|---------------|-----------|---------------|
| California    | 105000.0  | 73839.66      |
| Washington    | 102000.0  | 91397.85      |
| Massachusetts | 85000.0   | 62962.96      |
| New York      | 85000.0   | 57354.93      |
| Oregon        | 82846.0   | 63678.71      |

Bottom 5 sates

| Bottom 5 States | Paid Wage | Adjusted Wage |
|-----------------|-----------|---------------|
| Arkansas        | 63001.5   | 69308.58      |
| Oklahoma        | 60840.0   | 69215.02      |
| Montana         | 60000.0   | 59582.92      |
| Wyoming         | 58205.0   | 61723.22      |
| West Virginia   | 55000.0   | 60773.48      |

Then we can look at the updated list of top and bottom states which is:\
\
5 top:\

| Top 5 States | Adjusted Wage | Bottom 5 States | Adjusted Wage |
|--------------|---------------|-----------------|---------------|
| Washington   | 91397.85      | Rhode Island    | 59726.96      |
| Mississippi  | 82232.89      | Montana         | 59582.92      |
| Utah         | 80531.31      | New York        | 57354.93      |
| Kansas       | 78612.72      | Alaska          | 52731.71      |
| Tennessee    | 78292.13      | Hawaii          | 36646.66      |

```{r}

#  ADJUSTED_WAGE for all states
all_states_med_adjusted <- aggregate(salcost$ADJUSTED_WAGE, by = list(salcost$Work_State), FUN = median)
names(all_states_med_adjusted) <- c("State", "Median_Adjusted_Wage")

# Sort the all_states_med_adjusted data frame by median in descending order
all_states_med_adjusted <- all_states_med_adjusted[order(-all_states_med_adjusted$Median_Adjusted_Wage),]


#  PAID_WAGE for all states
all_states_med_paid <- aggregate(salcost$PAID_WAGE_PER_YEAR, by = list(salcost$Work_State), FUN = median)

names(all_states_med_paid) <- c("State", "Median_paid_Wage")

# Sort the all_states_med_paid data frame by median in descending order
all_states_med_paid <- all_states_med_paid[order(-all_states_med_paid$Median_paid_Wage),]


all_states_med <- merge(all_states_med_paid, all_states_med_paid, by = c("State"))

print(all_states_med)

```

-   **If we look at job subtitles, how does salary change for top and bottom 5 considering the cost of living?**

When examining job subtitles, we can observe significant changes in salaries based on the cost of living in the top and bottom five states. The top five states, which have the highest median salary, show a considerable decrease in wages after accounting for the cost of living. Conversely, the bottom five states, with the lowest median salary, exhibit a substantial increase in wages after adjusting for the cost of living, except for Montana and for management consultants in Oklahoma.

It's worth noting that, in the graph depicting the top states, the adjusted wage falls below \$50,000 for most business and data analysts, while these job titles in the bottom states remain above \$50,000.

The states with the most significant decrease in adjusted wages experience a decrease of approximately \$40,000 to \$50,000. However, not all job sub-titles exhibit an increase in adjusted wages; only business analysts and software engineers appear in all the states. The adjusted wage for business analysts indicates that low cost-of-living states provide better compensation than high cost-of-living states. Conversely, software engineers' paid wages show that they receive better compensation in high cost-of-living states, while adjusted wages demonstrate that they are compensated similarly across the board.

```{r}



###################################################

#Looking at the median change  of top 5 states by subtitle after adjusted wage
# First, extract the top 5 states
top_states <- head(all_states_med_paid, 5)$State

# Filter the data by the top 5 states
top_states_data <- salcost[salcost$Work_State %in% top_states, ]

# Calculate the median ADJUSTED_WAGE by job subtitle and state
state_job_medians <- aggregate(top_states_data$ADJUSTED_WAGE, 
                             by = list(top_states_data$Work_State, top_states_data$Job_Title_Sub),
                             FUN = median)

# Rename the columns of the state_job_means data frame
names(state_job_medians) <- c("State", "Job_Title_Sub", "Median_Adjusted_Wage")

# Order the data frame by state and mean adjusted wage in descending order
state_job_medians <- state_job_medians[order(state_job_medians$State, -state_job_medians$Median_Adjusted_Wage),]





###################comparison paid wage##################
########### comparison with payed wage  ################


# Calculate the mean PAID_WAGE_PER_YEAR by job subtitle and state
state_job_med_paidwage <- aggregate(top_states_data$PAID_WAGE_PER_YEAR, 
                             by = list(top_states_data$Work_State, top_states_data$Job_Title_Sub),
                             FUN = median)

# Rename the columns of the state_job_means data frame
names(state_job_med_paidwage) <- c("State", "Job_Title_Sub", "Med_Paid_Wage")

# Order the data frame by state and mean paid wage in descending order
state_job_med_paidwage <- state_job_med_paidwage[order(state_job_med_paidwage$State, -state_job_med_paidwage$Med_Paid_Wage),]




###############################

# Load ggplot2 library
library(ggplot2)
library(scales)

# Create a new data frame by merging state_job_means and state_job_means_paidwage data frames
state_job_means_all <- merge(state_job_medians, state_job_med_paidwage, by = c("State", "Job_Title_Sub"))



print(state_job_means_all)



##########We can also look at the bottom states:   


# First, extract the bottom 5 states
bottom_states <- tail(all_states_med_paid, 5)$State

# Filter the data by the bottom 5 states
bottom_states_data <- salcost[salcost$Work_State %in% bottom_states, ]

# Calculate the mean ADJUSTED_WAGE by job subtitle and state
state_job_med_b <- aggregate(bottom_states_data$ADJUSTED_WAGE, 
                             by = list(bottom_states_data$Work_State, bottom_states_data$Job_Title_Sub),
                             FUN = median)

# Rename the columns of the state_job_means data frame
names(state_job_med_b) <- c("State", "Job_Title_Sub", "Med_Adjusted_Wage")

# Order the data frame by state and mean adjusted wage in descending order
state_job_med_b <- state_job_med_b[order(state_job_med_b$State, -state_job_med_b$Med_Adjusted_Wage),]

# Print the state_job_means data frame
#print(state_job_med_b)

########################

# Calculate the median PAID_WAGE_PER_YEAR by job subtitle and state
state_job_med_paidwage_b <- aggregate(bottom_states_data$PAID_WAGE_PER_YEAR, 
                             by = list(bottom_states_data$Work_State, bottom_states_data$Job_Title_Sub),
                             FUN = median)

# Rename the columns of the state_job_means data frame
names(state_job_med_paidwage_b) <- c("State", "Job_Title_Sub", "Med_Paid_Wage")

# Order the data frame by state and mean paid wage in descending order
state_job_med_paidwage_b <- state_job_med_paidwage_b[order(state_job_med_paidwage_b$State, -state_job_med_paidwage_b$Med_Paid_Wage),]

# Print the state_job_means data frame
#print(state_job_med_paidwage_b)

############################### Here the colors are swaped. we have adjusted to be lighter and paid wage stronger color.. 

# Load ggplot2 library graph for bottom 5
library(ggplot2)
library(dplyr)

# Create a new data frame by merging state_job_means and state_job_means_paidwage data frames
state_job_means_all_b <- merge(state_job_med_b, state_job_med_paidwage_b, by = c("State", "Job_Title_Sub"))

print(state_job_means_all_b)



###################################new
library(ggplot2)
library(dplyr)

# Create the grouped bar plot
p1 <- ggplot(state_job_means_all, aes(x = State, y = Median_Adjusted_Wage, fill = Job_Title_Sub)) +
  geom_bar(position = "dodge", stat = "identity", width = 0.6) +
  geom_bar(aes(y = Med_Paid_Wage), position = "dodge", stat = "identity", width = 0.6, alpha = 0.5) +
  scale_fill_manual(values = c("#F8766D", "#00BA38", "#619CFF", "#FC8D62", "#8B00FF")) +
  labs(x = "State", y = "Med Salary", fill = "Job Title Sub") +
  ggtitle("Median Adjusted (darker) and Median Paid Wages (lighter)") +
  theme(plot.title = element_text(size = 16, hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma, limits = c(0, 150000))

# Create the grouped bar plot
p2 <- ggplot(state_job_means_all_b, aes(x = State, y = Med_Paid_Wage, fill = Job_Title_Sub)) +
  geom_bar(position = "dodge", stat = "identity", width = 0.6) +
  geom_bar(aes(y = Med_Adjusted_Wage), position = "dodge", stat = "identity", width = 0.6, alpha = 0.5) +
  scale_fill_manual(values = c("#F8766D", "#00BA38", "#619CFF", "#FC8D62", "#8B00FF")) +
  labs(x = "State", y = "Med Salary", fill = "Job Title Sub") +
  ggtitle("Median Adjusted (lighter) and Median Paid Wages (darker)") +
  theme(plot.title = element_text(size = 16, hjust = 0.5), axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma, limits = c(0, 150000))

# Display plots side-by-side with shared y-axis limits
library(gridExtra)
grid.arrange(p1, p2, ncol = 2, widths = c(1, 1))


```

\

**What are the states with the largest percentage change in adjusted wage?**

A look into the percentage change in Wages after cost of living is taken into account. We see that Hawaii has the highest negative percentage change with an adjusted salary of almost 48%, followed by New York, California, Massachusetts, and Oregon.\
On the other hand, the highest change in positive percentage change was Oregon with a 20% change followed by Kansas, Alabama, Oklahoma, and Georgia. **\
**

| State         | Positive % Change | Negative % Change |
|---------------|-------------------|-------------------|
| Mississippi   | 20.05             |                   |
| Kansas        | 15.61             |                   |
| Alabama       | 13.77             |                   |
| Oklahoma      | 13.77             |                   |
| Georgia       | 12.61             |                   |
| Oregon        |                   | -23.14            |
| Massachusetts |                   | -25.93            |
| California    |                   | -29.68            |
| New York      |                   | -32.52            |
| Hawaii        |                   | -48.27            |

```{r}
# Calculate the median cost index by state
state_cost_index <- aggregate(salcost$costIndex, by = list(salcost$Work_State), FUN = median)

# Rename the columns of the state_cost_index data frame
names(state_cost_index) <- c("State", "Median_Cost_Index")

# Order the data frame by mean cost index in descending order
state_cost_index <- data.frame(state_cost_index[order(-state_cost_index$Median_Cost_Index), ])

# Print the state_cost_index data frame
#print(state_cost_index)

#head(salcost)

```

A look into the percentage change in Wages after cost of living is taken into account. We see that Hawaii has the highest percentage change with an adjusted salary of almost 48%, followed by\

```{r}
# Calculate the mean paid wage per year by state
state_all_means_paid <- aggregate(PAID_WAGE_PER_YEAR ~ Work_State, data = salcost, FUN = mean)

# Rename the columns of the state_means data frame
names(state_all_means_paid) <- c("State", "Mean_Paid_Wage_Per_Year")

# Order the data frame by mean paid wage per year in descending order
state_all_means_paid <- state_all_means_paid[order(-state_all_means_paid$Mean_Paid_Wage_Per_Year),]

# Print the state_means data frame
#print(state_all_means_paid)
# Print the state medians data frame
#print(all_states_med_paid)


# Calculate the mean adjusted wage by state
state_all_means_adjusted <- aggregate(ADJUSTED_WAGE ~ Work_State, data = salcost, FUN = mean)

# Rename the columns of the state_means_adjusted data frame
names(state_all_means_adjusted) <- c("State", "Mean_Adjusted_Wage")

# Order the data frame by mean paid wage per year in descending order
state_all_means_adjusted <- state_all_means_adjusted[order(-state_all_means_adjusted$Mean_Adjusted_Wage),]

# Print the state_means data frame
#print(state_all_means_adjusted)
# Print the state medians data frame
#print(all_states_med_adjusted)

####################################################################
#Graph of all states means percentage change in wage after adjusted mean wage

# Merge the state_means and state_means_adjusted data frames by state
merged_means <- merge(state_all_means_paid, state_all_means_adjusted, by = "State")


# Add the WORK_STATE_ABBREVIATION column to merged_means
merged_means$WORK_STATE_ABBREVIATION <- salcost$WORK_STATE_ABBREVIATION[match(merged_means$State, salcost$Work_State)]

# Calculate the percentage change of the adjusted wage
merged_means$Percent_Change <- (merged_means$Mean_Adjusted_Wage - merged_means$Mean_Paid_Wage_Per_Year) / merged_means$Mean_Paid_Wage_Per_Year * 100

# Add a column indicating whether the percentage change is positive or negative
merged_means$Positive_Change <- ifelse(merged_means$Percent_Change > 0, "Yes", "No")

# Order the data frame by percent change in descending order
merged_means <- merged_means[order(-merged_means$Percent_Change),]

# Print the merged_means data frame
print(merged_means)


ggplot(data = merged_means, aes(x = WORK_STATE_ABBREVIATION)) +
  geom_bar(aes(y = Percent_Change, fill = Positive_Change), alpha = 0.8, stat = "identity") +
  scale_fill_manual(values = c("red", "green"), guide = "none" ) +
  geom_text(aes(y = Percent_Change, label = paste0(round(Percent_Change, 1), "%")), vjust = -0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Mean Adjusted Wage and Percent Change by State",
       x = "State Abbreviation",
       y = "Percent Change",
       color = "Positive Change") +
  theme_minimal()



```

```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE}
```
