---
title: "DS311 Project Q2"
author: "Gabrielle Salamanca"
date: "March 17, 2023"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## We-R-Finished

Before we can dive into the question, we must load in the necessary libraries and read the dataset into the markdown.

```{r}
# loading libraries
library(readxl)
library(caret)
library(ggplot2)
library(dplyr)

# dataset overview
salary <- read_excel("C:/Users/knigh/OneDrive/Desktop/Github/We-R-Finished/salary/salary_data_states.xlsx")

# renaming columns
colnames(salary)[1] = "Case Number"
colnames(salary)[2] = "Case Status"
colnames(salary)[3] = "Received Date"
colnames(salary)[4] = "Decision Date"
colnames(salary)[5] = "Employer Name"
colnames(salary)[6] = "Submitted Prevail Wage"
colnames(salary)[7] = "SPrW Unit"
colnames(salary)[8] = "Submitted Paid Wage"
colnames(salary)[9] = "SPaW Unit"
colnames(salary)[10] = "Job Title"
colnames(salary)[11] = "Work City"
colnames(salary)[12] = "Required Edu"
colnames(salary)[13] = "Required College Major"
colnames(salary)[14] = "Exp Req"
colnames(salary)[15] = "Exp Req (Months)"
colnames(salary)[16] = "Citizenship"
colnames(salary)[17] = "Prevail Wage SOC Code"
colnames(salary)[18] = "PWSOC Title"
colnames(salary)[19] = "Work State"
colnames(salary)[20] = "WS Abb"
colnames(salary)[21] = "WPostal Code"
colnames(salary)[22] = "Full Time"
colnames(salary)[23] = "Visa Class"
colnames(salary)[24] = "Prevail Wage/Yr"
colnames(salary)[25] = "Paid Wage/Yr"
colnames(salary)[26] = "Job Title Sub"
colnames(salary)[27] = "Order"
```

Now, we must remove the sub job titles that have teaching jobs. The question specifically asks for data-related salaries in the job sub-categories, therefore we must remove: professor, attorney, assistant professor, and teacher.

```{r}
# filter
sal <- salary %>%
  filter(!grepl("professor", `Job Title Sub`, ignore.case = TRUE) & 
           !grepl("attorney", `Job Title Sub`, ignore.case = TRUE) &
           !grepl("assistant professor", `Job Title Sub`, ignore.case = TRUE) & 
           !grepl("teacher", `Job Title Sub`, ignore.case = TRUE))
jobTitleSub <- as.factor(sal$`Job Title Sub`)
```

### II. What states (of those I am willing to move to) have the highest paying data-related salaries?

This is quite a broad range of states, but we will remove any US territories first and foremost for obvious reasons. This means Guam, Guamam, Palau, Northern Mariana Islands, Puerto Rico, and Virgin Islands will not be considered.

```{r}
# filtering states
datsal <- sal %>%
  filter(!grepl("Guam", `Work State`, ignore.case = TRUE) & 
           !grepl("Guamam", `Work State`, ignore.case = TRUE) &
           !grepl("Palau", `Work State`, ignore.case = TRUE) & 
           !grepl("Northern Mariana Islands", `Work State`, ignore.case = TRUE) &
           !grepl("Puerto Rico", `Work State`, ignore.case = TRUE) &
           !grepl("Virgin Islands", `Work State`, ignore.case = TRUE))
```

And we will make a new dataframe containing the states that had the largest frequency.

```{r}
topState <- c("Texas", "New York",
          	"New Jersey", "Illinois", "Massachusetts",
          	"Virginia", "Pennsylvania", "Washington",
          	"Michigan", "North Carolina")
ssal <- datsal[datsal$`Work State` %in% topState, ]
# data-related only
ssal$Sub <- factor(ssal$`Job Title Sub`, levels = c("business analyst", "data analyst", "data scientist", "management consultant", "software engineer"))
Sub <- ssal$Sub
```

#### b. Which companies have the highest salaries for those sub-types?

```{r}
# median salary
ssal %>%
  group_by(Sub, `Employer Name`) %>%
  summarise(med_salary = median(`Paid Wage/Yr`)) %>%
  arrange(Sub, desc(med_salary)) %>%
  group_by(Sub) %>%
  top_n(n = 5, wt = med_salary)
```

For each job sub type, we see the top 5 employers in terms of who has the highest median salary ceiling. Business analyst's highest ceiling is \$677508 with the University of Texas Administration, data analyst's is 185,000 with Knowledgent Group Inc, data scientist is with the PNC Services Group Inc at 205,000, management consultant at 450,000 with Baker Botts LLP, and software engineer at 1,224,801 with Abacus Technical Services LLC. Admittedly, these may be outliers, especially concerning software engineer. 

Now, how does this compare to the average salary? For the business analyst, the top 2 companies are unchanged: University of Texas Administration and Officemax Incorporated. Data analyst has stayed pretty consistent with its top 5, just shifting around the rankings. For data scientist, the first and last companies are unchanged. The management consultant salary is nearly the same, the last entry has changed. Finally, software engineer had the most change, only one company has stayed and has become the top rank. 

so we see that for best highest salary ceiling, the order is software engineer, business analyst, management consultant, data scientist, and data analyst. Management consultant, data scientist, and data analyst have stayed in the same place, while business analyst and software engineer have switched places depending the type of salary.

Let's try restricting the percentiles to see if this will change the top companies and the order.

```{r}
# median salary
ssal %>%
  filter(`Paid Wage/Yr` > quantile(`Paid Wage/Yr`, 0.05) & `Paid Wage/Yr` < quantile(`Paid Wage/Yr`, 0.95)) %>%
  group_by(Sub, `Employer Name`) %>%
  summarise(med_salary = median(`Paid Wage/Yr`)) %>%
  arrange(Sub, desc(med_salary)) %>%
  group_by(Sub) %>%
  top_n(n = 5, wt = med_salary)
```

We do end up with changes in the samples and different companies when we restrict the percentiles. In the max salary, business analyst, data analyst, and software engineer expanded; while data scientist and management consultant stayed the same. The order is software engineer, business analyst, data scientist, management consultant, and data analyst.

For the average salary, business analyst, management consultant, data analyst, and management consultant stayed the same sample size; and data scientist has greatly increased in that area. Business analyst, data scientist, and software have the same average: 124800. Data analyst falls short with 122000, as well as management consultant at 120000. 

We see that every single sub type went down in salary when we restricted the percentiles to 5th through 95th. Software engineer had the greatest loss out of all the job sub types, yet it still makes more out of all of them. This is quite confounding admittedly, how software engineer is such an umbrella term. But I suppose this means you do have more diverse skills, therefore could have a higher salary than most.

#### c. Will the answer change if I take standard of living into account?

If I took the standard of living into account, my answer will change. I would not want to move to New York, knowing how it's equally as expensive as the Bay Area. Texas suffers from extreme heat and cold, their infrastructure was unable to handle the extreme cold. Also those states on the east coast also suffer from severe winters, so I'd rather be somewhere where winter isn't as severe

I would probably, out of all the ten states, would want to move to Illinois or Washington. Despite knowing Illinois has a high crime rate, I personally have family in Illinois, so I would be okay moving there. Meanwhile, Washington perhaps is a little closer to what I'm used to, so it's also a safer bet for those who are non-white.



```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE}
```
